---
output: github_document
---

# RcTools

<!-- badges: start -->
<!-- badges: end -->

The goal of RcTools is to provide a collection of tools (functions) to be utilised by the Northern Three Report Cards' technical staff. These tools bridge the gap between: 

 1. Online data resources,
 2. Data analysis pipelines, and
 3. Publication ready figures and tables.

Specifically, the tools focus on tasks such as building commonly used spatial files, downloading datasets from APIs, calculating scores and grades, and producing stylised visual products such as plots, maps, and formatted xlsx files.

A website is available to fully explore what this package has to offer: [RcTools Website.](https://northern-3.github.io/RcTools/index.html)

# Getting Started

To install this package you will first need to download the [RTools package](https://cran.r-project.org/bin/windows/Rtools/rtools45/rtools.html) and install it on your computer (accept the defaults everywhere during the installation process).

Following this, you can install the development version of RcTools from [GitHub](https://github.com/) with:

```{r}
#| eval: FALSE

# install.packages("pak")
pak::pak("Northern-3/RcTools")

```

Finally, load the package just like you would any other R package:

```{r}
#| eval: FALSE

library(RcTools)

```

Once you have successfully loaded the package you are good to go!