---
title: "eReefs"
output: rmarkdown::html_vignette
format:
  html:
    code-overflow: wrap
vignette: >
  %\VignetteIndexEntry{eReefs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

eReefs has been a significant body of work for the Northern Three Partnership, with the ultimate objective being to obtain contextual information on the marine zone and river inputs across each region.

Accessing and using eReefs data presents one of the most significant hurdles to its adoption. Thus, a series of functions has been design specific for report card use that streamline and simplify the necessary aspects of the analysis. There are currently 7 exported functions conveniently split into three groups to discuss:

Data Extraction and Preparation Functions:

 - extract_ereefs()
 - ereefs_get_date_range()
 - ereefs_reproject()
 - ereefs_input_selection()

Data Visualisation Functions:

 - ereefs_map()
 - ereefs_dotplot()
 - ereefs_windrose()

# Data Extraction and Prepartion

All eReefs data is stored on THREDDS servers in the NetCDF format. It cannot be manually downloaded, and it is stored in a curvilinear format. Further, at the smallest scale (pre-cropping), the data is still several gb in size. Effectively, the data is painful to get a hold of.

At the most basic level, data can be extracted from eReefs as follows:

First a spatial boundary needs to be defined, in this case we are using a simple rectangle, but it could be any spatial object that you have.

```{r setup}
# echo: FALSE

#load libraries
library(RcTools)
library(tmap)

#define a spatial boundary
sf_obj <- system.file("extdata/boundary.gpkg", package = "RcTools")
sf_obj <- sf::st_read(sf_obj)

```

```{r}

#show object
tmap_mode("view")
tm_shape(sf_obj) + tm_borders()

```

Next, use the custom extract_ereefs() function.

```{r}
#| eval: FALSE

my_data <- extract_ereefs(
  Model = "https://dapds00.nci.org.au/thredds/dodsC/fx3/gbr4_bgc_GBR4_H2p0_B3p1_Cfur_Dnrt.ncml",
  Region = sf_obj,
  StartDate = "2022-03-01",
  EndDate = "2022-03-03",
  Variable = "Turbidity",
  Downsample = 0
)

```

```{r}
#| include: FALSE

#load in target data
load(system.file("extdata/turb_curvi.RData", package = "RcTools"))
my_data <- turb_curvi

```

Following this, data is converted from a curvilinear grid to a rectilinear grid. Note that is not always necessary (and sometimes fails), but it does make the data easier to save, understand, and work with.

```{r}

my_data_square <- ereefs_reproject(my_data)

```

And thats it. You now have an eReefs dataset on your computer.

Here is how to curvilinear object looks:

```{r}

tm_shape(my_data[,,,1]) + 
  tm_raster() +
  tm_shape(sf_obj) + 
  tm_borders()

```

and here is the regular grid:

```{r}

tm_shape(my_data_square[,,,1]) + 
  tm_raster() +
  tm_shape(sf_obj) + 
  tm_borders()

```

## Data Extraction in More Detail

Obviously things can get a bit more complicated than that, and a lot has been left unexplained. Below I try to provide a bit more detail.

First of all, the extract_ereefs() function has a few guardrails to be aware of:

 1. The variables that can be requested have been limited to only those that the Partnerships might find useful. Specifically, this extraction provides access to: "Turbidity", "Chlorophyll a", "DIN", "NH4", "NO3", "Secchi", "pH", "Wind", "True Colour". There are hundreds of other options.
 2. Datasets can be very, very, large. The extraction function provides the option to downsample data, that is to reduce the total number of cells extracted. The higher the number provided to the downsample arguement, the greater the reduction.
 3. There are a range of eReefs models you can access data from, with each of the models exploring a different scenario. The helper function ereefs_input_selection() assists with choosing the correct model and runs in the background when appropriate. (Note, above I just provided the direct link to skip this step).
 4. Every eReefs model spans a different period of time, some going back a decade, others only a few years, some recording up to 2023, others stopping in 2019. The function ereefs_get_date_range() can be used once you know what model you want, to determine the date range it offers.

Lets first look at picking a model, you will need to run this on your own machine as it is an interactive function:

```{r}
#| eval: FALSE

#to access this function you have to explicitly call it using the triple colon (:::), as it is not natively exported. This is because it is only supposed to be used within other eReefs functions
RcTools:::ereefs_input_selection()

```

Once selected the function returns the appropriate path. When this function is called within the main extract function the path is passed to the "model" argument.

::: {.callout-warning}
During the creation of this documentation (30/03/2026) it has been noted that the model options provided have been updated. Previously "eReefs GBR4 biogeochemistry and sediments v3.1 baseline catchment scenario" (option 5) was the preferred model to use. However, it has since been marked as "SUPERSEDED". It is likely that "eReefs GBR4 biogeochemistry and sediments v4.2 baseline catchment scenario" (option 11) will become the new default, however this model currently does not include recent data and appears to return NULL values often.
:::

Once you have your model, you can query its date range as follows:

```{r}

#query date range of the model. Note that if no model is provided, you are redirected to first choose a model
ereefs_get_date_range("https://dapds00.nci.org.au/thredds/dodsC/fx3/gbr4_bgc_GBR4_H2p0_B3p1_Cfur_Dnrt.ncml")

```

Finally, it is often the case that several variables are required within an area of interest, and/or several distinct dates are required. In this case, below is the code necessary to generalise the data extraction into a multiple extraction, simply create a matrix of input values and use pmap to iterate.

```{r}
#| eval: FALSE

#put arguments into objects
all_variables <- c("Turbidity", "Chl_a_sum")
all_start_dates <- "2022-03-01"
all_end_dates <- "2022-03-03"

#use expand grid to build a simple table of inputs
all_arguments <- expand.grid(all_variables, all_start_dates, all_end_dates, stringsAsFactors = FALSE)

#map over each argument, this returns a list of data extracts
multiple_extraction <- purrr::pmap(
  all_arguments,  
  function(Var1, Var2, Var3){
    extract_ereefs(
      Model = "https://dapds00.nci.org.au/thredds/dodsC/fx3/gbr4_bgc_GBR4_H2p0_B3p1_Cfur_Dnrt.ncml",
      Region = sf_obj, StartDate = Var2, EndDate = Var3, Variable = Var1)
    }
  )

#alternatively, you can manually build the table if you want more control (note that two different time spans are requested)
all_arguments_2 <- data.frame(
  "AllVariables" = c("Turbidity", "Chl_a_sum", "Turbidity", "Chl_a_sum"),
  "AllStartDates" = c("2022-01-01", "2022-01-01", "2023-01-01", "2023-01-01"),
  "AllEndDates" = c("2022-01-05", "2022-01-05", "2023-01-05", "2023-01-05")
)

#you can then reproject with the custom function using another map function
multiple_extraction <- purrr::map(multiple_extraction, ereefs_reproject)

```

This method returns a list of datasets, each object in the list is given an appropriate name based on the variable requested and its date range.

It is also important to note here that even if you are only interested in one variable (such as Chla), if you are requesting data that spans more than about a year it is a good idea to break the request into a list. Data requests that are too large can overload your computer.

## Saving a Data Extraction

Saving a data extraction from eReefs does not require any custom built functions. To save and read back in the files simply use the `stars::write_mdim()` and `stars::read_stars()` functions. When saving a single dataset this is achived as follows:

```{r}
#| eval: FALSE

#save
stars::write_mdim(my_data_square, "my_data_square.nc")

#read
my_data_square <- stars::read_stars("my_data_square.nc")

```

However, if you are unable to convert a curvilinear dataset into a regular grid, this saving method will not work. Instead you have to save the raw R object. This is obviously a subpar method of saving and locks the dataset to an R environment (hence the motivation to convert to a regular grid), however is sometimes unavoidable. In this case the method is as follows:

```{r}
#| eval: FALSE

#save
save(my_data_square, file = "my_data_square.RData")

#load (note that you do not assign the load to an object - it inherits its previous name from when it was saved).
load("my_data_square.RData")

```

Finally, it is often the case that several data objects are extracted from eReefs at the same time and returned as a list. In this case, the method can be vectorised using the same input table that was used to download the list of datasets (input table is from code a few chunks back up):

```{r}
#| eval: FALSE

#convert each row in the table into a file name
file_names <- purrr::map(1:nrow(all_arguments), \(x) paste(all_arguments[x, ], collapse = "_"))

#walk across file names and save each (hashed out to not save things unecessarily)
purrr::walk2(multiple_extraction, file_names, \(x, y) write_mdim(x, glue("{data_path}/{y}.nc")))

#build a list of files with the correct extension
files_to_read <- list.files(glue("{data_path}"))[str_detect(list.files(glue("{data_path}")), ".nc")]

#read in all files
mutiple_load <- purrr::map(files_to_read, \(x) read_stars(glue("{data_path}/{x}")))

```

# Data Visualisation

There are two key ways that eReefs data is currently visualised in the Technical Reports. These are as dotplots, and as maps. In both cases custom functions have been devised to assist in their creation.

## Plotting

Plotting eReefs data is fairly simple, the only variables you need to control are the label names and if you want a log y-axis:

```{r}

#load in target data
nc <- stars::read_mdim(system.file("extdata/turb_reg.nc", package = "RcTools"))

#run the function
ereefs_dotplot(nc) 

```

This function accepts both a list of data objects, or a single object. This is again to address the scenario in which multiple requests are made to  eReefs and a list of datasets is returned. You should be careful to only provide a list that is A) of a single variable, and B) with a chronological order (i.e. don't provide a list with datasets from the same time period).

While this function is quite simple, it should be noted that this is a double edge sword. For example, if you want to focus on a particular timeframe you will need to edit the data before plotting. If this is of interest I highly reccomend learning about [stars objects](https://r-spatial.github.io/stars/articles/stars1.html).

In the rare case that you were interested in creating a windrose, that option is also available. Obviously you would have to have downloaded wind data though:

```{r}

#load in target data
nc <- stars::read_mdim(system.file("extdata/wind_reg.nc", package = "RcTools"))

#run the function
suppressWarnings(ereefs_windrose(nc, Aggregation = "Annual"))

```

## Mapping

Mapping eReefs data is a bit more complicated as there are several variations of maps packed into the one function. Specifically, you can create:

 - concentration maps (e.g. for Chla),
 - true colour maps (e.g. to get water colour), or
 - vector field maps (to show wind direction).

In each case, the minimum required inputs are fairly similar, you simply need to supply the data and define the map type. As long as you have not massively edited the data provided by the extraction function (for example by changing variable names), then the function should happily accept each type of data. 

Examples of how to call each map type are presented below, and the true colour map is visualised.

```{r}
#| eval: FALSE

tmap_mode("plot")

#load in target data
nc <- stars::read_mdim(system.file("extdata/turb_reg.nc", package = "RcTools"))

#run the function
ereefs_map(nc, MapType = "Concentration")

#load in target data
nc <- stars::read_mdim(system.file("extdata/wind_reg.nc", package = "RcTools"))

#run the function
ereefs_map(nc, MapType = "Vector Field")

#load in target data
nc <- stars::read_mdim(system.file("extdata/tc_reg.nc", package = "RcTools"))

#run the function
ereefs_map(nc, MapType = "True Colour")

```

```{r}
#| echo: FALSE

tmap_mode("plot")
nc <- stars::read_mdim(system.file("extdata/tc_reg.nc", package = "RcTools"))

```

```{r}
#| echo: FALSE
#| output: TRUE

#run the function
ereefs_map(nc, MapType = "True Colour")

```

This function also accepts both a list of data objects, or a single object. This is again to address the scenario in which multiple requests are made to  eReefs and a list of datasets is returned.

If you only provide the dataset and the map type, the mapping function will assume some reasonable defaults. For example you can see that in this case the maps have been provided with a monthly aggregation. You can define how much to aggregate the data using the "Aggregation" argument and pick between "Month", "Season", "Financial", "Annual".

## Extending mapping

You may notice that the maps presented are not the most stylish, for example there are no region borders to indicate where the data should end, and there are no basin outlines to related data to specific locations on land. This function actually produces a tmap object, not a map directly. What this means is you can further use the tmap package to add additional layers and style the map however you want.

A basic example of further styling is presented below, and importantly please understand that the order of the layers in the code is the order in which they are placed on the map (first layer = lowest layer on map, second layer is placed on top, etc.). It is recommended that the R package [Tmap](https://r-tmap.github.io/tmap/) is reviewed to further understanding mapping using this package.

```{r}

#run the function again, but this time save as an object and aggregate to annual
my_map <- ereefs_map(nc, MapType = "True Colour", Aggregation = "Annual")

#use the standard tmap structure to build extract layers around our data
my_map + #note how the object is added here, it could be added at any point in this call including in between layers or after layers
  tm_shape(sf_obj) + 
  tm_borders()
  
```