---
title: "Report Card Scoring"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Report_Card_Scoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

A key component of the Northern Three Report Cards is calculating scores and grades for a range of indicator. To streamline this process three key functions have been developed that work in conjuction:

 1. value_to_score()
 2. score_to_grade()
 3. save_n3_table()

Additional helper functions have been written that support these three key functions, which include:

 - bind_letter_to_score()
 - define_colour_scheme()
 - conditional_formatter()
 - weight_scores()

# General Workflow

The general workflow of these functions can be demonstrated as follows. Take a defined/loaded/calculated dataframe such as this one:

```{r}

library(RcTools)

df <- data.frame(
  Basin = c("Black", "Ross", "Haughton", "Suttor"),
  Indicator = c("DIN", "DIN", "Low_DO", "Low_DO"),
  Value = c(0.002, 0.017, 87.6, 104),
  Wqo = c(0.02, 0.02, 90, 90),
  Sf = c(0.38, 0.38, 70, 70),
  Eightieth = c(0.0154, 0.0154, 101.12, 101.12),
  Twentieth = c(0.0026, 0.0026, 75.84, 75.84)
)

```

First run the value_to_score() function. This function takes several inputs depending on the type of value and the way you wish to score it. In this first example, we are scoring using the freshwater water quality method, and therefore are required to supply a large amount of information:

```{r}

#run the function on our example data
df <- df |> 
  value_to_score(
    value = Value,
    value_type = "Water Quality",
    water_type = "Freshwater",
    indicator = Indicator,
    wqo = Wqo,
    sf = Sf,
    eightieth = Eightieth,
    twentieth = Twentieth
  )

```

This function returns the original table with an additional column added. The column inherits the name of the value column that was provided, plus the string "Score" after it. For example if the value column was named "MyValue", the new column would be named "MyValueScore".

Next the score_to_grade function can be run. In this example we will continue using the dataframe from above. However, this function can take any number of score columns as inputs and will return the same number of new columns with the associated text grade.

```{r}

#run the function on our example data
df <- score_to_grade(df, ValueScore)

```

```{r}
#| eval: FALSE

#if you wanted to get grades for multiple columns this can be acheived as follows
x <- score_to_grade(x, c(ScoreA, ScoreB, ScoreC))

```

The column(s) inherits the name of the score column that was provided, plus the string "Grade" after it. Further, if the string "Score" is detected, it is replaced with "Grade". For example if the value column was named "MyValue", the new column would be named "MyValueGrade", or if the column was named "MyValueScore" then the new column would be named "MyValueGrade"

Finally, a table that has been scored and potentially graded can then be saved using the save_n3_table() function. This function is unique in that it will apply excel native conditional formatting to the outputs of the function. I.e., you can save a dataframe with scores, and automatically apply the required colour scaling (red to green).
